---
title: "Metaheurystyki i Technologie Wieloagentowe - Lab3"
author: "Bartłomiej Twardowski, B.Twardowski@ii.pw.edu.pl"
date: "1/2/2015"
output: ioslides_presentation
---

## Organizational information

### e-learning part

```{r,eval=FALSE}
e_learning_part <- function(student) {
 if(student.missing_classes > 1) {
   TRUE
 } else {
    FALSE // ;-) 
 }
}
```

## Lab3 - Exercise

### Data preparation
### Simple analysis
### Predict feature with metaheuristic

## Data preparation

### Go to Polish Stock Exchange website:

http://www.gpwinfostrefa.pl/GPWIS2/pl/quotes/archive/1

### Download historical data 

Choose one year period (e.g. from *2010/01/01* to *2011/01/01*)
and download data for selected company (e.g. *AMICA*).

Data will be downloaded in Excel file named like: _PLAMICA00010.xls_.

**TODO:** Open Excel and look at the downloaded data.

## Load data to R

To load data from Excel to R we can use package _xlsx_.

```{r,eval=FALSE}
install.packages('xlsx')
```

Reading data into R:

```{r,echo=FALSE}
require('xlsx')
stock_data <- read.xlsx('../data//PLAMICA00010.xls', 1)
# Date type conversion
stock_data$Data=as.Date(stock_data$Data,format="%Y-%m-%d")
```

Look what's inside dataframe:
```{r,eval=FALSE}
summary(stock_data)
head(stock_data)
```

## Analysis - Plot the data

```{r}
par(mfrow=c(2,2))
plot(stock_data$Kurs.zamkniecia~stock_data$Data, xlab="Data", ylab="Closing price (PLN)")
plot(stock_data$Kurs.zamkniecia~stock_data$Kurs.otwarcia, xlab="Starting price (PLN)", ylab="Closing price (PLN)")
plot(stock_data$Kurs.zamkniecia~stock_data$Wolumen.obrotu, xlab="Trading volume", ylab="Closing price (PLN)")
plot(stock_data$Kurs.zamkniecia~stock_data$Kurs.maksymalny, xlab="Maximla price (PLN)", ylab="Closing price (PLN)")
````

## Aggregate data by week

```{r}
stock.week.avg <- as.numeric(
    aggregate(
      stock_data$Kurs.zamkniecia,
      by=list(floor((stock_data$Data-stock_data$Data[1])/7)),
      FUN=mean)$x
  )
```

## Plot the result

```{r}
par(mfrow=c(1,2))
with(stock_data, {
  plot(Kurs.zamkniecia~Data,type='l',xlab="Date",ylab="Closing price(PLN)")
})
plot(stock.week.avg,type='l',xlab="Week",ylab="Avg closing price(PLN)")
```


## Predicting avarage week price

Predicting week price base on waighted avarage from last weeks.
Parameters of prediction are *weights* for avarage.

```{r}
predict_price <- function(weights=c(w1,w2,w3,w4), weekNum, weekAvg) {
  prediction <- 0
  count <-0
  for(i in 1:length(weights)) {
    if(weekNum>i) {
      prediction <- prediction + weights[i]*weekAvg[weekNum-i] 
      count <- count + 1
    }
  }
  return (prediction/count)
}
```

Simple avarage from last for weeks:

```{r}
predict_price(c(1,1,1,1),5,stock.week.avg)
```

## Parameters evealuation 

Cost function will be Mean Root Square function calculated on
all dataset for given parameters (weights for our prediction).

```{r}
predict_eval <- function(weights=c(w1,w2,w3,w4), weekAvg){
  predictions <-as.numeric(lapply(4:length(weekAvg),
                    FUN=function(w) predict_price(weights,w,weekAvg)
                  )
                )
  
  ## data normalization!
  max_val=max(abs(c(weekAvg[-(1:3)],predictions)))
  return(sqrt(mean((weekAvg[-(1:3)]/max_val-predictions/max_val)^2)))
}
```

## Finding best weights

To find weights for our prediction which best fits the model we us one of
metaheuristic algorithm: **Particle Swarm Optimization** from package
**pso**.

```{r,eval=FALSE}
install.packages('pso')
?pso::psoptim
```

## Runing optimizer

```{r}
library(pso)
#random initial weights
init_weights=runif(4,0,1)
opt_par<-psoptim(init_weights, predict_eval, gr=NULL, stock.week.avg,
                 lower=rep(0,length(init_weights)), 
                 upper=rep(1,length(init_weights)),
                 control=list(maxit=100,type='SPSO2011',reltol=1e-8,vectorize=TRUE))

opt_par$par #best weights
opt_par$value #minimal cost(eval) function
```

## Result visualization

```{r}
test_prediction <-as.numeric(lapply(
  2:length(stock.week.avg),
  FUN=function(w) predict_price(opt_par$par,w,stock.week.avg)
  ))

par(mfrow=c(1,1))
plot(test_prediction,type='o',col='red',ylim=c(min(test_prediction),max(test_prediction)),xlab="Week",ylab="Avg price(PLN)")
points(test_prediction[-(1)],type='o')
```






